Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-04-15 2020/RSD_GPU/RSD2020.tex
Поправлены таблицы

\section{Multidimensional parallel global search algorithm}
Consider the problem of finding the global minimum of the \(N\)-dimensional function \(\varphi(y)\) in the hyperinterval  \(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\).  We will assume that the function satisfies the Lipschitz condition with an a priori unknown constant \(L\).
\begin{equation}
\label{task}
\varphi(y^*)=\min\{\varphi(y):y\in D\},
\end{equation}
\begin{equation}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty.
\end{equation}

@@

The adaptive dimensionality reduction scheme changes the order in which subproblems are solved: they will be solved not one by one (in accordance with their hierarchy in the task tree), but simultaneously, i.e., there will be a number of subproblems that are in the process of being solved. Under the new scheme:	
	\begin{itemize}
		\item to calculate the value of the function of the \(i\)-th level from (\ref{subTasks}), a new \(i+1\)-th level problem is generated in which trials are carried out, after which a new generated problem is included in the set of existing problems to be solved;
		\item the iteration of the global search consists in choosing \(p\) (the most promising) problem from the set of existing problems in which trials are carried out; points for new trials are determined in accordance with the algorithm of subsection 1;
		\item the minimum values of functions from (\ref{subTasks}) are their current estimates based on the accumulated search information.
	\end{itemize}

A brief description of the main steps of a block adaptive dimensionality reduction scheme is as follows.
Let the nested subproblems in the form (\ref{subTasks}) be solved using the global search algorithm described in Section 2. Then each subproblem (\ref{subTasks}) can be assigned a numerical value called the characteristic of this problem. As such, we can take the maximum characteristic \(R(t)\) of the intervals formed in this problem. In accordance with the rule for calculating characteristics, the higher the value of the characteristic, the more promising the subproblem is in the continued search for the global minimum of the original problem (\ref{task}).

@@

\ref{fig:Parallel_Computing} shows the general scheme of organizing calculations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\varphi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using CPU only. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

@@

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time span for calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems \cite{Modorskii2016}, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series).

@@

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Jones2009}, DIRECT\textit{l} \cite{Gablonsky2001} and Global Search Algorithm (GSA) from Section 2.  A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Gaviano2003}. The global minimum $y*$ was considered found if the algorithm generated a trial point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\| \leqslant \delta $.  Also, the size of the neighborhood was chosen (in accordance with \cite{Sergeyev2006}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved , a and b are the boundaries  of the search domain $D$, the parameter  $\Delta ={{10}^{-6}}$ for N=4 and $\Delta ={{10}^{-7}}$ for $N=5$. When using the GSA method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class $r=5.6$; the parameter for constructing the Peano curve was fixed, $m=10$. The maximum number of iterations allowed was $K_{max} = 10^6$.

@@

As can be seen from Table \ref{table:average_iters}, the sequential GSA surpasses the DIRECT and DIRECT\textit{l} methods in all classes of problems in terms of the average number of iterations. At the same time, in the 5-Hard class, none of the methods solved all the problems: DIRECT failed to solve 16 problems, DIRECTl and GSA -- 4 problems each.

%\begin{table}
                %\caption{Speedup on CPU}
                %\label{table:average_AS_CPU}
                %\center
                %\begin{tabular}{lccccccccccc}
                               %%\hline\noalign{\smallskip}
																	 %& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               %\noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                                   %& \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               %\noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               %& Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               %\noalign{\smallskip} \hline \noalign{\smallskip}                               
                               %$p=2$&   3,49 & 2,35 &   & 0,86 & 2,03 &   & 1,71 & 1,17 &   & 0,95 & 0,4 \\
                               %$p=4$&   8,22 & 7,65 &   & 6,74 & 8,56 &   & 2,01 & 1,9  &   & 1,85 & 1,53 \\
                               %\noalign{\smallskip}\hline
                %\end{tabular}
%\end{table}

@@

\begin{table}
                \caption{Speedup on CPU}
                \label{table:average_MMAS_CPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																	 & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                                   & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=2$&    6,62  & 3,23  &   & 2,13 & 6,07  &   & 3,22 & 1,48 &   & 0,66 & 1,93 \\
                               $p=4$&    21,46 &  9,97 &   & 6,9  & 16,27 &   & 5,23 & 2,42 &   & 0,85 & 1,9 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on one GPU}
                \label{table:average_AS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    5,06  & 3,93  &   & 1,19 	 & 5,41  &   & 2,2	    & 1,92	 &   & 0,47	& 2,21 \\
                               $p=256$ &    19,91 & 14,95 &   & 11,9 	 & 23,36 &   & 8,28	& 6,91	     &   & 3,4	    & 6,1 \\
							   $p=1024$&    15,62 & 52,91 &   & 22,12	 & 55,22 &   & 5,21	& 19,95      &   & 2,18	& 4,82 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on two GPUs}
                \label{table:average_MMAS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,32	 & 6,83	 &   & 8,19	     & 10,66  &   & 2,44	 & 1,63	 &   & 1,33	 & 1,58 \\
                               $p=256$ &    38,96	 & 31,52 &   & 25,43	 & 48,23  &   & 8,01	 & 7,04	 &   & 2,31	 & 3,42 \\
							   $p=1024$&    128,39	 & 83,46 &   & 98,43	 & 146,23 &   & 20,45	 & 11,06 &   & 2,75	 & 2,48 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

@@

\begin{table}
                \caption{Speedup on six GPUs}
                \label{table:GKLS6_average_MMAS_GPU}
                \center
                \begin{tabular}{lccc}
                               %\hline\noalign{\smallskip}
                &  Iteration speedup & &   Time speedup \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,32	 & & 6,83	  \\
                               $p=256$ &    38,96	 & & 31,52  \\
															 $p=1024$&    128,39	 & & 83,46  \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

>>>>

\section{Multidimensional parallel global search algorithm}\label{Sec_GSA}
Consider the problem of finding the global minimum of the \(N\)-dimensional function \(\varphi(y)\) in the hyperinterval  \(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\).  We will assume that the function satisfies the Lipschitz condition with an a priori unknown constant \(L\).
\begin{equation}
\label{task}
\varphi(y^*)=\min\{\varphi(y):y\in D\},
\end{equation}
\begin{equation}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty.
\end{equation}

@@

The adaptive dimensionality reduction scheme changes the order in which subproblems are solved: they will be solved not one by one (in accordance with their hierarchy in the task tree), but simultaneously, i.e., there will be a number of subproblems that are in the process of being solved. Under the new scheme:	
	\begin{itemize}
		\item to calculate the value of the function of the \(i\)-th level from (\ref{subTasks}), a new \(i+1\)-th level problem is generated in which trials are carried out, after which a new generated problem is included in the set of existing problems to be solved;
		\item the iteration of the global search consists in choosing \(p\) (the most promising) problem from the set of existing problems in which trials are carried out; points for new trials are determined in accordance with the parallel global search algorithm from Section \ref{Sec_GSA};
		\item the minimum values of functions from (\ref{subTasks}) are their current estimates based on the accumulated search information.
	\end{itemize}

A brief description of the main steps of a block adaptive dimensionality reduction scheme is as follows.
Let the nested subproblems in the form (\ref{subTasks}) be solved using the global search algorithm described in Section \ref{Sec_GSA}. Then each subproblem (\ref{subTasks}) can be assigned a numerical value called the characteristic of this problem. As such, we can take the maximum characteristic \(R(t)\) of the intervals formed in this problem. In accordance with the rule for calculating characteristics, the higher the value of the characteristic, the more promising the subproblem is in the continued search for the global minimum of the original problem (\ref{task}).

@@

Fig. \ref{fig:Parallel_Computing} shows the general scheme of organizing computations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\varphi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using CPU only. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

@@

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time of calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems \cite{Modorskii2016}, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series).

@@

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Jones2009}, DIRECT\textit{l} \cite{Gablonsky2001} and Global Search Algorithm (GSA) from Section 2.  A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Gaviano2003}. The global minimum $y^*$ was considered found if the algorithm generated a trial point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\| \leqslant \delta $.  The size of the neighborhood was chosen (in accordance with \cite{Sergeyev2006}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved , $a$ and $b$ are the boundaries  of the search domain $D$, the parameter  $\Delta ={{10}^{-6}}$ for $N=4$ and $\Delta ={{10}^{-7}}$ for $N=5$. When using the GSA method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class $r=5.6$; the parameter for constructing the Peano curve was $m=10$. The maximum number of iterations allowed was $K_{max} = 10^6$.

@@

As can be seen from Table \ref{table:average_iters}, the sequential GSA surpasses DIRECT and DIRECT\textit{l} methods in all classes of problems in terms of the average number of iterations. At the same time, in the 5-Hard class, none of the methods solved all the problems: DIRECT failed to solve 16 problems, DIRECTl and GSA -- 4 problems each.

%\begin{table}
                %\caption{Speedup on CPU}
                %\label{table:average_AS_CPU}
                %\center
                %\begin{tabular}{lccccccccccc}
                               %%\hline\noalign{\smallskip}
																	 %& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               %\noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                                   %& \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               %\noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               %& Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               %\noalign{\smallskip} \hline \noalign{\smallskip}                               
                               %$p=2$&   3,5 & 2,4 &   & 0,9 & 2,0 &   & 1,7 & 1,2 &   & 1,0 & 0,4 \\
                               %$p=4$&   8,2 & 7,7 &   & 6,7 & 8,6 &   & 2,0 & 1,9  &   & 1,9 & 1,5 \\
                               %\noalign{\smallskip}\hline
                %\end{tabular}
%\end{table}

@@

\begin{table}
                \caption{Speedup on CPU}
                \label{table:average_MMAS_CPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																	 & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                                   & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=2$&    6,6  & 3,2   &   & 2,1  & 6,6  &   & 3,2 & 1,5 &   & 0,7 & 2,1 \\
                               $p=4$&    21,5 &  10,0 &   & 6,9  & 19,2 &   & 5,2 & 2,4 &   & 0,9 & 2,3 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on one GPU}
                \label{table:average_AS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
																& \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    5,1  & 3,9  &   & 1,2 	 & 9,3   &   & 2,2	& 1,9	     &   & 0,5	& 4,0  \\
                               $p=256$ &    19,9 & 15,0 &   & 11,9 	 & 39,6  &   & 8,3	& 6,9	     &   & 3,4	& 11,0 \\
							   $p=1024$&    15,6 & 52,9 &   & 22,1	 & 105,7 &   & 5,2	& 20,0      &   & 2,2	& 10,2 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

\begin{table}
                \caption{Speedup on two GPUs}
                \label{table:average_MMAS_GPU}
                \center
                \begin{tabular}{lccccccccccc}
                               %\hline\noalign{\smallskip}
                & \multicolumn{5}{c}{ Iteration speedup} & & \multicolumn{5}{c}{ Time speedup}\\
                               \noalign{\smallskip} \cline{2-6} \cline{8-12}  \\
                               & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} &  & \multicolumn{2}{c}{ $N=4$ } & & \multicolumn{2}{c}{$N=5$} \\
                               \noalign{\smallskip} \cline{2-3} \cline{5-6}  \cline{8-9} \cline{11-12} \noalign{\smallskip}
                               & Simple & Hard & & Simple & Hard &  & Simple & Hard & & Simple & Hard  \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,3	 & 6,8	 &   & 8,2	 & 19,8  &   & 2,4	 & 1,6	 &   & 1,3	 & 3,2 \\
                               $p=256$ &    39,0	 & 31,5 &   & 25,4	 & 57,1  &   & 8,0	 & 7,0	 &   & 2,3	 & 2,4 \\
							   $p=1024$&    128,4	 & 83,5 &   & 98,4	 & 267,9 &   & 20,4  & 11,0  &   & 2,8	 & 5,2 \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}

@@

\begin{table}
                \caption{Speedup on six GPUs}
                \label{table:GKLS6_average_MMAS_GPU}
                \center
                \begin{tabular}{lccc}
                               %\hline\noalign{\smallskip}
                &  Iteration speedup & &   Time speedup \\
                               \noalign{\smallskip} \hline \noalign{\smallskip}                               
                               $p=64$  &    11,3	 & & 6,8	  \\
                               $p=256$ &    39,0	 & & 31,5  \\
															 $p=1024$&    128,4	 & & 83,5  \\
                               \noalign{\smallskip}\hline
                \end{tabular}
\end{table}
